\proglang{Stan}: A Probabilistic Programming Language

### 2014-03-01 src/docs/papers/stan-language/article.tex
added corrections based on comments from Aki

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.  As of version 2.2.0,
  \proglang{Stan} provides full Bayesian inference for continuous-variable
  models through Markov chain Monte Carlo methods such as
  the No-U-Turn sampler, an adaptive form of Hamiltonian Monte
  Carlo sampling.  Penalized maximum likelihood estimates are
  calculated using optimization methods such as the
  Broyden–Fletcher–Goldfarb–Shanno algorithm.

@@

The probability distribution defined by a \proglang{Stan} program is
intended to have unconstrained support (i.e., no points of zero
probabilty), which greatly simplifies the task of writing samplers or
optimizers.  To achieve unbounded support, variables declared with
constrained support are transformed to an unconstrained space.  For
instance, variables declared on $[0,1]$ are log-odds transformed and
non-negative variables declared to fall in $[0,\infty)$ are log
transformed.  More complex transforms are required for simplexes (a
reverse stick-breaking transform) and covariance and correlation
matrices (Cholesky factorization).  The dimensionality of the
resulting probabilty function may change as a result of the
transform. For example, a $K \times K$ covariance matrix requires only
${K \choose 2} + K$ unconstrained parameters, and a $K$-simplex
requires only $K-1$ unconstrained parameters.

The unconstrained parameters over which the model is defined are
inverse transformed back to their constrained forms before executing
the model code.  To account for the change of variables, the log
absolute Jacobian determinant of the inverse transform is added to the
overall log probabilty function.%
%
\footnote{For optimization, the Jacobian adjustment is suppressed to
  guarantee the optimizer finds the maximum of the log probability
  function on the constrained parameters.  The calculation of the
  Jacobian is controlled by a template parameter in the \proglang{C++}
  code generated for a model.}
%
The gradients of the log probabilty function exposed include the
Jacobian term.

@@

The list in Figure~\ref{variable-kinds.fig} is completed with two
types of local variables, loop indices and traditional local
variables.  Unlike \proglang{BUGS}, \proglang{Stan} allows local
variables to be declared and assigned.  For example, it is possible to
compute the sum of the squares of entries in an array or vector
\code{y} as follows.%
%
\footnote{The built-in squared norm function is the right way to
  implement sums of squares in \proglang{Stan}.}
%
\begin{Code}
{ 
  real sum_of_squares;
  sum <- 0;
  for (n in 1:N)
    sum <- sum + y[n] * y[n];
}  
\end{Code}
%
Local variables may not be declared with constraints, because there is
no location at which it makes sense to test that they satisfy the
constraints.

@@

The following reparameterization employs a non-centered
parameterization for comparison.
%
\begin{Code}
parameters {
  vector[K] beta_raw;    
}
transformed parameters {
  vector[K] beta;
  beta <- mu + sigma * beta_raw;
}
model {
  beta ~ normal(0,1);
}
\end{Code}
%
This reparameterization removes the particular correlations amongst the
hierarchical variables that would otherwise limit the effectiveness
of the samplers.  We recommend starting with the more natural centered
parameterization and moving to the non-centered parameterization if
sampling of the coefficients does not mix well.%
%
\footnote{As the saying goes in computer science, it is easier to
  optimize a correct program than debug an optimized program.}

@@

Vectors are declared with their sizes and matrices with their number
of rows and columns.

@@

@@

The default optimizer in \proglang{Stan} is the
Broyden-Fletcher-Goldfarb-Shannon-Boyd (BFGS) optimizer.  BFGS is a
quasi-Newton optimizer that evaluates gradients directly, then uses
the gradients to update an approximation to the Hessian.  Plans are in
the works to also include the more involved, but more scalable
limited-memory BFGS (L-BFGS) scheme.  \citet{NocedalWright:2006} cover
both BFGS and L-BFGS samplers.

@@

>>>>

%% an abstract and keywords
\Abstract{ \proglang{Stan} is a probabilistic programming language for
  specifying statistical models. A \proglang{Stan} program
  imperatively defines a log probability function over parameters
  conditioned on specified data and constants.  As of version 2.2.0,
  \proglang{Stan} provides full Bayesian inference for continuous-variable
  models through Markov chain Monte Carlo methods such as
  the No-U-Turn sampler, an adaptive form of Hamiltonian Monte
  Carlo sampling.  Penalized maximum likelihood estimates are
  calculated using optimization methods such as the
  Broyden-Fletcher-Goldfarb-Shanno algorithm.

@@

The probability distribution defined by a \proglang{Stan} program is
intended to have unconstrained support (i.e., no points of zero
probability), which greatly simplifies the task of writing samplers or
optimizers.  To achieve unbounded support, variables declared with
constrained support are transformed to an unconstrained space.  For
instance, variables declared on $[0,1]$ are log-odds transformed and
non-negative variables declared to fall in $[0,\infty)$ are log
transformed.  More complex transforms are required for simplexes (a
reverse stick-breaking transform) and covariance and correlation
matrices (Cholesky factorization).  The dimensionality of the
resulting probability function may change as a result of the
transform. For example, a $K \times K$ covariance matrix requires only
${K \choose 2} + K$ unconstrained parameters, and a $K$-simplex
requires only $K-1$ unconstrained parameters.

The unconstrained parameters over which the model is defined are
inverse transformed back to their constrained forms before executing
the model code.  To account for the change of variables, the log
absolute Jacobian determinant of the inverse transform is added to the
overall log probability function.%
%
\footnote{For optimization, the Jacobian adjustment is suppressed to
  guarantee the optimizer finds the maximum of the log probability
  function on the constrained parameters.  The calculation of the
  Jacobian is controlled by a template parameter in the \proglang{C++}
  code generated for a model.}
%
The gradients of the log probability function exposed include the
Jacobian term.

@@

The list in Figure~\ref{variable-kinds.fig} is completed with two
types of local variables, loop indices and traditional local
variables.  Unlike \proglang{BUGS}, \proglang{Stan} allows local
variables to be declared and assigned.  For example, it is possible to
compute the sum of the squares of entries in an array or vector
\code{y} as follows.%
%
\footnote{The built-in squared norm function is the right way to
  implement sums of squares in \proglang{Stan}.}
%
\begin{Code}
{ 
  real sum_of_squares;
  sum_of_squares <- 0;
  for (n in 1:N)
    sum_of_squares <- sum_of_squares + y[n] * y[n];
}  
\end{Code}
%
Local variables may not be declared with constraints, because there is
no location at which it makes sense to test that they satisfy the
constraints.

@@

The following reparameterization employs a non-centered
parameterization for comparison.
%
\begin{Code}
parameters {
  ...
  vector[K] beta_raw;  // replaces beta as parameter  
}
transformed parameters {
  vector[K] beta;      // beta now a transformed parameter
  beta <- mu + sigma * beta_raw;
}
model {
  beta_raw ~ normal(0,1);  // beta now sampled from unit normal
  ...
}
\end{Code}
%
This reparameterization removes the particular correlations amongst the
hierarchical variables that would otherwise limit the effectiveness
of the samplers.  We recommend starting with the more natural centered
parameterization and moving to the non-centered parameterization if
sampling of the coefficients does not mix well.%
%
\footnote{As the saying goes in computer science, it is easier to
  optimize a correct program than debug an optimized program.}

@@

Vectors are declared with their sizes and matrices with their number
of rows and columns.  Vector, row vector, and matrix elements are
accessed using bracket notation, as in \code{y[3]} for the third
element of a vector or row vector and \code{a[2,3]} for the element in
the third column of the second row of a matrix.  Indexing begins from
1.

The notation \code{a[2]} accesses the second row of matrix \code{a}.
Rows, columns and subblocks of matrices can be accessed using
\code{row}, \code{col}, and \code{block} functions.  Slices of arrays
can be accessed using the \code{head}, \code{tail}, and \code{segment}
functions.  There are also special functions for creating a diagonal
matrix from a vector and accessing the diagonal of a vector.

@@

The manual contains a chapter discussing the efficiency tradeoffs and
motivations for separating arrays and matrices.

@@

The default optimizer in \proglang{Stan} is the
Broyden-Fletcher-Goldfarb-Shanno (BFGS) optimizer.  BFGS is a
quasi-Newton optimizer that evaluates gradients directly, then uses
the gradients to update an approximation to the Hessian.  Plans are in
the works to also include the more involved, but more scalable
limited-memory BFGS (L-BFGS) scheme.  \citet{NocedalWright:2006} cover
both BFGS and L-BFGS samplers.

@@

We'd like to thank Aki Vehtari for comments and corrections on a draft
of the paper.
